\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{natbib}

\titlespacing*{\section}{0pt}{1.2ex plus 0.5ex minus 0.2ex}{0.8ex plus 0.2ex}
\titlespacing*{\subsection}{0pt}{0.8ex plus 0.3ex minus 0.1ex}{0.5ex plus 0.1ex}

\setlength{\parskip}{0.4em}
\setlength{\parindent}{0pt}

\title{\textbf{Cuffless Blood Pressure Estimation from Photoplethysmography\\Using Time-Series Feature Engineering and Ensemble Learning}}
\author{Vignan Kamarthi \and Ariv Ahuja}
\date{DS4400: Machine Learning 1 -- Spring 2026}

\begin{document}
\maketitle

\section{Problem Description}

Cardiovascular diseases (CVDs) are the leading cause of mortality worldwide. Blood pressure (BP) is a critical CVD risk indicator, yet conventional measurement requires an inflatable cuff, a method that is intrusive and only captures isolated snapshots rather than continuous readings. This makes it unsuitable for ongoing monitoring throughout daily life. Cuffless BP monitoring from wearable sensors would enable early hypertension detection and real-time cardiovascular tracking.

Photoplethysmography (PPG), the optical pulse-sensing technology found in consumer smartwatches, captures blood volume changes that correlate with arterial pressure dynamics. This project frames cuffless BP estimation as a \textbf{supervised regression task}: given time-series features extracted from a 10-second PPG waveform segment, predict the corresponding systolic blood pressure (SBP, the peak pressure when the heart contracts) and diastolic blood pressure (DBP, the lowest pressure when the heart relaxes), both measured in mmHg.

The core methodological contribution is a \textbf{dual feature extraction pipeline} combining Catch22 canonical time-series features with entropy-based complexity measures, followed by ensemble learning for BP regression. The Catch22 approach adapts methods from prior work on physiological signal classification \citep{boda2025ai4pain}, while the entropy-based features build on established information-theoretic measures including permutation entropy \citep{bandt2002permutation} and complexity-entropy analysis \citep{rosso2007distinguishing}, applied in ongoing biosignal complexity research. Together, these capture complementary aspects of PPG morphology relevant to arterial pressure.

\section{Dataset}

This project uses \textbf{PulseDB v2.0} \citep{wang2023pulsedb}, a large-scale photoplethysmography dataset publicly available at \url{https://github.com/pulselabteam/PulseDB}. PulseDB aggregates recordings from two independent hospital systems: the MIMIC-III Waveform Database Matched Subset (Beth Israel Deaconess Medical Center, Boston) and VitalDB (Seoul National University Hospital, South Korea), which provides geographic and demographic diversity. The dataset contains 5,245,454 ten-second segments from 5,361 subjects, totaling approximately 14,570 hours of recording. Each segment consists of a raw PPG waveform sampled at 125 Hz (1,250 data points), paired with ground-truth SBP and DBP labels derived beat-by-beat from invasive arterial blood pressure (ABP) waveform recordings.

PulseDB provides three evaluation partitions: a training set (2,506 subjects), a calibration-based test set (held-out segments from the same training subjects), and a calibration-free test set (279 completely disjoint subjects never seen during training). The calibration-free split is the more clinically meaningful evaluation, as it tests whether the model generalizes to entirely new individuals without any patient-specific calibration data.

\section{Approach and Methodology}

\subsection{Feature Extraction}
Each 10-second raw PPG segment (1,250 samples at 125 Hz) is transformed into a fixed-length feature vector through three complementary extraction methods:
\begin{enumerate}[nosep]
    \item \textbf{Catch22} (22 features): Canonical time-series features capturing autocorrelation structure, distributional properties, successive differences, and fluctuation scaling \citep{lubba2019catch22}. Applied following prior work on physiological signal classification \citep{boda2025ai4pain}.
    \item \textbf{Statistical} ($\sim$8 features): Mean, median, standard deviation, skewness, kurtosis, root mean square, min, max.
    \item \textbf{Entropy-based} ($\sim$5 features): Sample entropy, permutation entropy \citep{bandt2002permutation}, approximate entropy, spectral entropy, and Hjorth complexity. These quantify signal regularity and complexity, capturing aspects of PPG morphology not represented by Catch22 or statistical summaries.
\end{enumerate}

The resulting feature vector ($\sim$35 features per segment) is compact enough for efficient training over the full 5.2 million segments while capturing distinct dimensions of the PPG waveform relevant to blood pressure.

\subsection{Feature Normalization}
Standard scaling (z-score normalization) fitted on training data only, then applied to test data to prevent data leakage. Normalization is critical for Ridge regression, which is sensitive to feature scale. Tree-based models (Random Forest, XGBoost, LightGBM) are scale-invariant since they split on thresholds rather than feature magnitudes, but we normalize uniformly across all models for pipeline consistency and will run an ablation to confirm the impact.

\subsection{Models}
Separate models trained for SBP and DBP. Hyperparameters tuned via cross-validation.

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Model} & \textbf{Type} & \textbf{Role} \\
\midrule
Ridge Regression & Regularized linear (L2 penalty) & Interpretable baseline \\
Random Forest & Bagging ensemble & Variance reduction; feature importance \\
XGBoost & Gradient boosting & Sequential error correction \\
LightGBM & Histogram-based boosting & Efficient large-scale training \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Language and Packages}
Python: \texttt{pycatch22}, \texttt{antropy}, \texttt{EntropyHub}, \texttt{scikit-learn}, \texttt{xgboost}, \texttt{lightgbm}. Feature extraction parallelized on NEU Explorer cluster (SLURM, 56 CPUs).

\subsection{Evaluation Metrics}
Model performance is assessed against the Association for the Advancement of Medical Instrumentation (AAMI) clinical standard, which requires mean absolute error (MAE) below 5 mmHg and standard deviation of error (SD) below 8 mmHg. MAE serves as the primary accuracy metric, while SD captures prediction consistency. Root mean squared error (RMSE) complements MAE by penalizing large errors more heavily, surfacing dangerous outlier predictions that MAE alone may mask. $R^2$ quantifies the proportion of blood pressure variance explained by the model. Finally, Bland-Altman analysis provides a clinical agreement plot of mean predicted value against prediction difference, visualizing systematic bias and limits of agreement as is standard in medical device validation. All metrics are reported on both the calibration-based and calibration-free test sets to quantify the generalization gap.

\section{Outcome}

The primary objective is to predict SBP and DBP from PPG waveforms with MAE approaching the AAMI clinical standard of 5 mmHg. Beyond raw accuracy, this project aims to demonstrate that the combination of Catch22 canonical time-series features and entropy-based complexity measures captures BP-relevant PPG physiology effectively enough for clinical-grade estimation. We will quantify the calibration-free versus calibration-based performance gap to characterize how well the models generalize to unseen subjects, and conduct feature importance analysis and ablation studies to identify which PPG characteristics are most predictive of arterial pressure.

\section{Plan (Rough, Subject to Change)}

\begin{table}[h]
\centering
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Task} & \textbf{Owner} \\
\midrule
Dataset acquisition and EDA & Vignan \\
Feature extraction pipeline & Vignan \\
Baseline models (Ridge, RF) & Ariv \\
Boosted models + tuning & Vignan \\
Milestone report & Both \\
Feature ablation study & Ariv \\
Bland-Altman + AAMI evaluation & Vignan \\
Final report + presentation & Both \\
\bottomrule
\end{tabular}
\end{table}

\newpage
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
